Learning with Misspecified Models: The Case of Overestimation\\ \href{https://jimena-galindo.github.io/fileRepo/draft.pdf}{\normalsize Click Here for the Latest Version}

### 2023-10-27 Paper/draft.tex
draft edits

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=2.5cm]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

\setlength{\parindent}{25pt}
\setlength{\parskip}{0pt}
\setlength{\skip\footins}{0.25cm} % margin before footnotes
\definecolor{myblue}{RGB}{31, 61, 122}
\definecolor{mypink}{RGB}{204, 0, 82}
\newtheorem{assumption}{Assumption}
\newtheorem{proposition}{Proposition} % Use the 'proposition' counter for numbering
\usepackage{rotating, graphicx}

\newcommand{\EE}[1]{\mathbb{E}\left[#1\right]}
\newcommand{\PP}[1]{\mathbb{P}\left(#1\right)}
\newcommand{\CP}[2]{\mathbb{P}\left(#1 \,| \, #2 \right)}
\newcommand{\CE}[2]{\mathbb{E}\left[#1\,|\,#2 \right]}
\usepackage[bottom]{footmisc}
\usepackage[doublespacing]{setspace}
\usepackage[normal]{caption}
\usepackage{dsfont}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{hyperref}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Learning with Mental Models: the case of overestimation},
  pdfauthor={Jimena Galindo},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Learning with Mental Models: the case of overestimation}
\author{Jimena Galindo}
\date{October 24, 2023}

@@

As an example consider a student who needs to decide how much effort to
put into studying for an exam. Their decision will depend on two
factors: their belief about their intrinsic ability, and their belief
about how difficult the exam will be. The outcome they observe will be
affected by how much they decide to study. Imagine that the student puts
in a moderate amount of effort and gets a surprisingly good grade. Did
they get a good grade because they are smarter than they thought? Or
because the exam was easier than they had anticipated? Their future exam
preparation strategy will depend fundamentally on which line of
reasoning they take. This feedback loop is referred to as an
\emph{endogenous information 
process} and is at the center of the forces I study.

To understand what the main forces at play are, I compare a set of
theories that model learning in settings with endogenous information,
and which can rationalize the persistence of misspecified beliefs. I
develop a unifying framework that nests multiple theories of learning
and generates testable predictions for each of them. Then I cast this
framework in a laboratory experiment and test the predictions to
identify which of the theories are consistent with the behavior I
observe in the lab. The experiment features an agent who needs to learn
two parameters: one that pertains their own characteristics (an
\emph{ego-relevant} parameter), and an exogenously determined state---in
the context of the student preparing for exams, the two parameters would
be their intrinsic ability and the difficulty of the exam.

Misspecified beliefs about an ego-relevant parameter are often referred
to as overconfidence (or underconfidence) and have been documented by
behavioral scientists\footnote{\citet{kelley1980} 
provides a review of the psychology literature.} and
economists\footnote{See \citet{benjamin2019} 
for a review of the literature in economics.} in a variety of settings.
\citet{Oster2013} find that subjects who are at high risk of having
Huntington's disease overestimate their probability of being healthy and
make retirement decisions as if they were healthy. \citet{Hoffman2020}
show that workers overestimate the quality of their match to their
current employment and are unlikely to look for other opportunities.
\citet{Camerer1999} find that entrepreneurs are overconfident about the
quality of their enterprise, which leads to excessive entry and early
exit from markets.

In all of these examples, holding an incorrect belief about a
fundamental leads to suboptimal choices with potentially high costs. In
spite of the abundance of evidence, the scope of the existing research
is limited in terms of the frameworks it considers. Most of the
experimental evidence documenting the bias is collected in settings
where subjects are passive learners.\footnote{
\citet{Gotte2022} and \citet{Ozyilmaz2022} are exceptions that study settings with endogenous information processes.}
They observe a noisy signal and report their beliefs over subsequent
rounds.\footnote{\citet{Bracha2012} and 
\citet{Mobius2022} are some examples.} Although these studies provide
important insights, they are not flexible enough to incorporate the
richer theories that have been proposed more recently. In particular,
they do not allow the study of endogenous learning or for learning about
multiple parameters at once.\footnote{\citet{Coutts2020} 
studies an environment with an ego-relevant parameter and an exogenous state but does not 
incorporate the endogenous information process.}

@@

When the setting features learning traps, even an agent who incorporates
all information correctly, may fall into learning traps as outlined by
\citet{Hestermann2021}. These traps are characterized by a combination
of an incorrect belief and an optimal action which produce information
that confirms the incorrect belief. Once an agent falls into a trap, the
belief will be stable and even with a correctly specified model of
learning, they will not be able to abandon their misspecified beliefs.
If the agent is dogmatic about their initial belief,
\citet{Heidhues2018} show that they will inevitably fall into a trap and
thus will be able to rationalize and sustain their initial
misspecification
belief.\footnote{\citet{Gotte2022} study the case of agents with dogmatic initial beliefs in a 
laboratory experiment.}

\citet{Ba2023} moves away from dogmatism and endows the agents with a
mechanism through which they can abandon incorrect beliefs. This allows
them to avoid falling into learning traps. To do so agents perform
Bayesian hypothesis tests that evaluate which is the more likely
parameter out of two possibilities. By doing this, she characterizes the
set of situations in which, even agents who consider alternative
paradigms, may become trapped.\footnote{A 
similar mechanism is proposed by \citet{Schwarstein2021} in a setting with persuasion.}

Lastly, misattribution bias is the more classical explanation and has
been widely studied in behavioral
science.\footnote{See \citet{kelley1980} for a review.} Agents who
suffer from misattribution bias will attribute successes to their own
ability---the ego-relevant parameter---and failures to bad luck---the
state. Under this model of learning, even an agent who initially has a
correct initial belief may become overconfident if they observe a
sequence of successes. In this case, the main driver of the bias is not
an initial misspecification or the presence of learning traps, it is the
updating procedure
itself.\footnote{A more general framework that can be used to model 
this bias has also been proposed by \citet{Brunnermeier2005} and empirically studied by \citet{Bracha2012}.}

These theories provide the main building blocks for a simplified
framework that can be directly implemented in a laboratory experiment.
In the experiment subjects make choices and receive feedback that
depends on their own ability, an exogenous parameter and the choice they
made. I track their choices as well as their beliefs about their own
ability. The goal is to identify which of the 3 forces---the presence of
learning traps, misspecified initial beliefs, or misatribution
bias---better explains the observed behavior. To determine the fit of
the models, I compare the behavior predicted by each theory to the
benchmark given by the fully Bayesian updating procedure.

I also study whether the learning mechanism is inherently linked to the
ego-relevance of the parameters or if it is a more general phenomenon. I
vary the degree of ego-relevance by introducing a treatment in which
subjects learn about the ability of another participant. In this
treatment, the participants know only the gender and nationality of the
other, and thus can induce stereotypes---a different type of
misspecification.\\
If correct learning about the parameters happens at higher rates in the
stereotype treatment, it would suggest that the bias is intrinsically
linked to the ego-relevance of the parameter. In contrast, if similar
biased behavior arises in both treatments, it is more likely that the
main driver of these types of misspecified beliefs is the updating
procedure itself or the endogenous information process.

Although some agents do fall into learning traps, I find that the
behavior of most subjects is better explained by misattribution bias:
good news are treated as signaling high ability, while bad news are
attributed to a low state. I also find that misattribution is no more
prevalent in the ego-relevant than in the stereotype condition. This
suggests that the main driver of the misspecification is the updating
procedure; however, the underlying mechanism by which the bias is
generated may be different in both treatments: While in the ego-relevant
condition subjects prefer to hold themselves in high esteem, in the
stereotype condition updating seems to be driven by some sort of bias
overcorrection---when subjects realize that they underestimated the
ability of another participant based on their gender and nationality,
they compensate by overestimating their ability.

Finally, I estimate the structural parameters of the models to study
model-heterogeneity in the sample. I find that even at an individual
level the behavior is better explained by a general model of
misattribution bias for most subjects. There is a smaller group of
subjects that can be better explained through dogmatic beliefs and
hypothesis testing and none of them behave in line with the fully
Bayesian benchmark.

In what follows I first discuss the theoretical framework and the
predictions of each of the theories. Then I introduce a unifying example
and my hypotheses. In section 4 I describe the experimental design and
in section 5 I present the data and the results. Section 6 outlines the
estimation of the parameters and the model fit analysis.

@@

An agent is of type \(\theta \in \Theta\) and faces an unknown exogenous
state \(\omega\) drawn from some density \(f\) over \(\Omega\). The
agent knows the distribution of \(\omega\) but not its realized value.
His belief about the state, \(p_0(\omega)\), coincide with the true
distribution, \(f\); and his belief about the type is \(p_0(\theta)\).
Let the agent's true type be \(\theta^{*}\), and the realized state be
\(\omega^{*}\).

@@

The agent chooses an action \(a\in A\) and observes a noisy outcome
\(h\). The outcome is a function of the agent's type, the state, and the
action. In particular \(h = h(\theta^*, \omega^*, a) + \varepsilon\)
with \(h(\cdot)\) increasing in both \(\theta^*\) and \(\omega^*\), and
such that conditional on a pair of parameters \((\theta, \omega)\),
there is a unique optimal action. Additionally,
\(\varepsilon\sim N(0, \sigma)\) is noise in the output.

After observing the outcome, the agent updates his beliefs about
\(\theta\) and \(\omega\) using some algorithm and moves on to the next
period. He repeats this process infinitely many times. I make the
simplifying assumption that the agent is myopic and chooses the action
that maximizes the payoff in each period. This assumption simplifies the
analysis and plays a role in whether an agent who updates their beliefs
using Bayes rule would learn the truth or not, however, for the main
theory discussed in this section, the results hold even when relaxing
this assumption.

A key notion in this setting is that of a self-defeating
equilibrium\footnote{This notion is an adaptation of the Berk-Nash equilibrium in \citet{Esponda2016} to this setting with only one agent}.
A \emph{self-defeating equilibrium} is a belief and action pair,
\((p, a)\), such that the agent's belief about their type is
misspecified, and the outcome generated by the action is consistent with
the misspecified belief. The agent's belief is said to be \emph{stable}
when this happens and the average output is

\[\bar{h}(a^*) = h(\hat{\theta}, \hat{\omega}, a)\] Where
\(\hat{\theta}\) and \(\hat{\omega}\) is the type-state pair on which
the misspecified belief places all probability. That is, the agents
belief is \(p(\hat{\theta}, \hat{\omega}) = 1\) and
\(p(\theta, \omega) = 0\) for all other values of the parameter.

Within this framework, I consider two nested theories of belief
updating: the first one is a dogmatic modeler adapted from
\citet{Heidhues2018},The second one is a switcher, as in \citet{Ba2023}.
The dogmatic modeler can be seen as a switcher with an infinitely sticky
initial belief so the two theories are nested. However, outside of the
limiting case, they produce different predictions about the agent's
behavior. I discuss each of them separately in what follows.

@@

A dogmatic agent is characterized by a learning model where he does not
update their beliefs about \(\theta\), instead, he holds a degenerate
belief that places all probability on \(\hat{\theta}\), which is
potentially misspecified. In this case, no matter how much information
he gathers against being of type \(\hat{\theta}\), he will not update
his beliefs about it. Any discrepancies between the observed outcomes
and his believed type are incorporated using the Bayes rule to update
his beliefs about \(\omega\). \citet{Heidhues2018} show that, under
certain assumptions on the per-period
utility,\footnote{The assumptions are that $u$ is twice continuously differentiable with: (i)$u_{aa}<0$ and $u_a(\underline{a},  \theta, \omega)>0>u_a(\Bar{a}, \theta, \omega)$, (ii) $u_{\theta}, u_{\omega}>0$ and (iii) $u_{a\theta}<0$ and $u_{a\omega}>0$. Where $\underline{a}$ is the
minimal action and $\Bar{a}$ is the maximal one. The direction of the derivatives is a normalization
and the results would hold even when the signs are reversed.} a dogmatic
modeler will inevitably fall into a self-defeating equilibrium. The
equilibrium will be such that the outcomes they observe reinforce their
belief on \(\omega\) in such a way that as \(t\to\infty\) the agent will
be sure that the state is some \(\omega^{'}\) consistent with their
believed type and the observed data. In other words, they will be in a
self-defeating equilibrium with a stable belief that places probability
one on the incorrect parameters \((\hat{\theta}, \omega_{\infty})\)

The mechanism by which the dogmatic agent falls into the self-defeating
equilibrium is the following: Suppose the agent holds the misspecified
belief that they are type \(\hat{\theta}>\theta^*\). For any prior over
\(\omega\), the agent will be disappointed by the outcome. That is:
\[\bar{h}(\theta^*, \omega^*, a_t^*) := \frac{1}{T}\sum_t \left[h(\theta^*, \omega^*, a_t^*)+\varepsilon_t\right]<\frac{1}{T}\sum_t \left[h(\hat{\theta}, \mathbb{E}(\omega), a_t^*)+\varepsilon_t\right]\]
Where the right-hand-side of the inequality represents the average
outcome that the agent expected.

There are two possible sources for the disappointment: the first is that
the realized state is lower than the expected state; the second source
is that the agent is of type \(\theta^*\) and therefore, for all
possible states, his gain will be lower than what he expected.

Because the agent is dogmatic, he will not update his beliefs about
\(\theta\) and as a consequence will attribute the disappointment to the
state being lower than expected. He will continue to update in this way
until he converges to a belief about \(\omega\) that is stable. Such a
belief will explain the observed utility perfectly and allow the agent
to rationalize his dogmatic belief about \(\theta\). Under the
assumptions of \citet{Heidhues2018}, there is a unique value of
\(\omega\) at which the belief is stable, I will refer to such value as
\(\omega_\infty\). This mechanism is illustrated in Example 1.

\textbf{Example 1: } \emph{Set \(A = \Omega=[\underline{a}, \infty)\)
and consider a student with intrinsic ability \(\theta^*\geq 0\) who
faces a grading procedure \(\omega^*\) that is unknown to them. The
student knows that a higher \(\omega^*\) is more likely to yield a
higher grade. In particular, she knows the grade is given by
\((\theta^*+a)\omega^*\).}

\emph{The student must choose an effort level \(a\), which determines
her grade. For whatever the chosen effort level, the student must pay a
cost \(c(a) = \frac{1}{2}a^2\) and observes a grade of
\((\theta^*+a)\omega^*\). She repeats this process for infinitely many
periods. Assume also that the student's prior is such that
\(\mathbb{E}[\omega]= \omega^*\) and she is dogmatic about being of type
\(\hat{\theta}>\theta^*\).\footnote{The example is illustrated for an overconfident student but 
the results are symmetric for a dogmatic student who initially 
places probability one on some $\tilde{\theta}<\theta^*$.} Therefore,
the student's payoff in period \(t\) is given by }

@@

In particular, for an agent that starts with a dogmatic belief that they
are of type \(\hat{\theta}\) but is willing to consider the alternative
explanation that they are of type \(\tilde{\theta}\), the agent will
switch to the alternative if:

@@

In order to compare the predictions of the theories discussed above, I
develop a unifying example where the forces behind each of the theories
can be isolated. The example is a modified from \citet{Heidhues2018} and
is adapted to be implementable in the lab.

@@

\caption{Probability of success for each type, gamble and effort level}
\label{tab:DGP}
\end{table}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=2.5cm]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

\setlength{\parindent}{25pt}
\setlength{\parskip}{0pt}
\setlength{\skip\footins}{0.25cm} % margin before footnotes
\definecolor{myblue}{RGB}{31, 61, 122}
\definecolor{mypink}{RGB}{50, 107, 31}
\newtheorem{assumption}{Assumption}
\newtheorem{proposition}{Proposition} % Use the 'proposition' counter for numbering
\usepackage{rotating, graphicx}

\newcommand{\EE}[1]{\mathbb{E}\left[#1\right]}
\newcommand{\PP}[1]{\mathbb{P}\left(#1\right)}
\newcommand{\CP}[2]{\mathbb{P}\left(#1 \,| \, #2 \right)}
\newcommand{\CE}[2]{\mathbb{E}\left[#1\,|\,#2 \right]}
\usepackage[bottom]{footmisc}
\usepackage[doublespacing]{setspace}
\usepackage[normal]{caption}
\usepackage{dsfont}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{hyperref}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdfauthor={Jimena Galindo},
  colorlinks=true,
  linkcolor={mypink},
  filecolor={myblue},
  citecolor={mypink},
  urlcolor={mypink},
  pdfcreator={LaTeX via pandoc}}

\title{Learning with Misspecified Models: The Case of Overestimation\\
\href{https://jimena-galindo.github.io/fileRepo/draft.pdf}{\normalsize Click Here for the Latest Version}}
\author{Jimena Galindo}
\date{October 26, 2023}

@@

As an example, consider a student who needs to decide how much effort to
put into studying for an exam. Their decision will depend on two
factors: their belief about their intrinsic ability and their belief
about how difficult the exam will be. The outcome they observe will be
affected by how much they decide to study. Imagine that the student
exerts a moderate amount of effort and gets a surprisingly good grade.
Did they get a good grade because they are smarter than they thought? Or
because the exam was easier than they had anticipated? Their future
exam-preparation strategy will depend fundamentally on which line of
reasoning they take. This feedback loop is referred to as an
\emph{endogenous information process} and is at the center of the forces
I study.

To understand what main forces are at play, I compare a set of theories
that model learning in settings with endogenous information and can
rationalize the persistence of misspecified beliefs. I develop a
unifying framework that nests multiple theories of learning and
generates testable predictions for each of them. Then, I cast this
framework in a laboratory experiment and test the predictions to
identify which of the theories are consistent with the behavior I
observe in the lab. The experiment features an agent who needs to learn
two parameters: one that pertains to their own characteristics (an
\emph{ego-relevant} parameter), and an exogenously determined state. In
the context of the student preparing for exams, the two parameters would
be their intrinsic ability and the difficulty of the exam.

Misspecified beliefs about an ego-relevant parameter are often referred
to as overconfidence (or underconfidence) and have been documented by
behavioral
scientists\footnote{\citet{kelley1980} provides a review of the psychology literature.}
and
economists\footnote{See \citet{benjamin2019} for a review of the literature in economics.}
in a variety of settings. \citet{Oster2013} find subjects who are at
high risk of having Huntington's disease overestimate their probability
of being healthy, and they make retirement decisions as if they were
healthy. \citet{Hoffman2020} show workers overestimate the quality of
their match to their current employment and are unlikely to look for
other opportunities. \citet{Camerer1999} find entrepreneurs are
overconfident about the quality of their enterprise, which leads to
excessive entry and early exit from markets.

In all of these examples, holding an incorrect belief about a
fundamental leads to suboptimal choices with potentially high costs. In
spite of the abundance of evidence, the scope of the existing research
is limited in terms of the frameworks it considers. Most of the
experimental evidence documenting the bias is collected in settings
where subjects are passive
learners.\footnote{\citet{Gotte2022} and \citet{Ozyilmaz2022} are exceptions that study settings with endogenous information processes.}
They observe a noisy signal and report their beliefs over subsequent
rounds.\footnote{\citet{Bracha2012} and \citet{Mobius2022} are some examples.}
Although these studies provide important insights, they are not flexible
enough to incorporate the richer theories that have been proposed more
recently. In particular, they do not allow the study of endogenous
learning or for learning about multiple parameters at
once.\footnote{\citet{Coutts2020} studies an environment with an ego-relevant parameter and an exogenous state but does not 
incorporate the endogenous information process.}

@@

When the setting features learning traps, even an agent who incorporates
all information correctly may fall into learning traps, as outlined by
\citet{Hestermann2021}. These traps are characterized by a combination
of an incorrect belief and an optimal action that produce information
that confirms the incorrect belief. Once an agent falls into a trap, the
belief will be stable and even with a correctly specified model of
learning, they will not be able to abandon their misspecified beliefs.
If the agent is dogmatic about their initial belief,
\citet{Heidhues2018} show they will inevitably fall into a trap and thus
will be able to rationalize and sustain their initial misspecification
belief.\footnote{\citet{Gotte2022} study the case of agents with dogmatic initial beliefs in a 
laboratory experiment.}

\citet{Ba2023} moves away from dogmatism and endows the agents with a
mechanism through which they can abandon incorrect beliefs. This allows
them to avoid falling into learning traps. To do so, agents perform
Bayesian hypothesis tests that evaluate which of two possible parameters
is more likely. By doing so, she characterizes the set of situations in
which even agents who consider alternative paradigms may become
trapped.\footnote{A similar mechanism is proposed by \citet{Schwarstein2021} in a setting with persuasion.}

Lastly, misattribution bias is the more classical explanation and has
been widely studied in behavioral
science.\footnote{See \citet{kelley1980} for a review.} Agents who
suffer from misattribution bias will attribute successes to their own
ability---the ego-relevant parameter---and their failures to bad
luck---the state. Under this model of learning, even an agent who
initially has a correct initial belief may become overconfident if they
observe a sequence of successes. In this case, the main driver of the
bias is not an initial misspecification or the presence of learning
traps; it is the updating procedure
itself.\footnote{A more general framework that can be used to model this bias has also been proposed by \citet{Brunnermeier2005} and empirically studied by \citet{Bracha2012}.}

These theories provide the main building blocks for a simplified
framework that can be directly implemented in a laboratory experiment.
In the experiment subjects make choices and receive feedback that
depends on their own ability, an exogenous parameter and the choice they
made. I track their choices as well as their beliefs about their own
ability. The goal is to identify which of the three forces---the
presence of learning traps, misspecified initial beliefs, or
misatribution bias---better explains the observed behavior. To determine
the fit of the models, I compare the behavior predicted by each theory
with the benchmark given by the fully-Bayesian updating procedure.

I also study whether the learning mechanism is inherently linked to the
ego-relevance of the parameters or if it is a more general phenomenon. I
vary the degree of ego-relevance by introducing a treatment in which
subjects learn about the ability of another participant. In this
treatment, the participants know only the gender and nationality of the
other, and thus can induce stereotypes---a different type of
misspecification. If correct learning about the parameters happens at
higher rates in the stereotype treatment, this finding would suggest the
bias is intrinsically linked to the ego-relevance of the parameter. By
contrast, if similar biased behavior arises in both treatments, the main
driver of these types of misspecified beliefs is more likely to be the
updating procedure itself or the endogenous information process.

Although some agents do fall into learning traps, I find the behavior of
most subjects is better explained by misattribution bias: good news are
treated as signaling high ability, whereas bad news are attributed to a
low state. I also find misattribution is no more prevalent in the
ego-relevant than in the stereotype condition. This finding suggests the
main driver of the misspecification is the updating procedure; however,
the underlying mechanism by which the bias is generated may be different
in both treatments. Whereas subjects in the ego-relevant condition
prefer to hold themselves in high esteem, in the stereotype condition,
updating seems to be driven by some sort of bias overcorrection---when
subjects realize they underestimated the ability of another participant
based on their gender and nationality, they compensate by overestimating
their ability.

Finally, I estimate the structural parameters of the models to study
model-heterogeneity in the sample. I find that even at an individual
level, the behavior is better explained by an average model of
misattribution bias for most subjects. A smaller group of subjects
exists that can be better explained through dogmatic beliefs and
hypothesis testing, and very few of them behave in line with the fully
Bayesian benchmark.

In what follows, I first discuss the theoretical framework and the
predictions of each of the theories. Then, I introduce a unifying
example and my hypotheses. In section 4, I describe the experimental
design, and in section 5, I present the data and the results. Section 6
outlines the estimation of the parameters and the model fit analysis.

@@

An agent of type \(\theta \in \Theta\) faces an unknown exogenous state
\(\omega\) drawn from density \(f\) over \(\Omega\). The agent knows the
distribution of \(\omega\) but not its realized value. His belief about
the state \(\omega\in\Omega\) is \(p_0(\omega)\) and coincides with the
true distribution, \(f\). His belief about the type \(\theta\in\Theta\)
is \(p_0(\theta)\). Let the agent's true type be \(\theta^{*}\), and the
realized state be \(\omega^{*}\).

@@

The agent chooses an action \(a\in A\) and observes a noisy outcome
\(h\). The outcome is a function of the agent's type, the state, and the
action. In particular, \(h = h(\theta^*, \omega^*, a) + \varepsilon\)
with \(h(\cdot)\) increasing in both \(\theta^*\) and \(\omega^*\), and
such that conditional on a pair of parameters \((\theta, \omega)\),
there is a unique optimal action. Additionally,
\(\varepsilon\sim N(0, \sigma)\) is noise in the outcome.

After observing the outcome, the agent updates his beliefs about
\(\theta\) and \(\omega\) using some algorithm and moves on to the next
period. He repeats this process infinitely many times. I make the
simplifying assumption that the agent is myopic and chooses the action
that maximizes the payoff in each period. This assumption simplifies the
analysis and plays a role in whether an agent who updates his beliefs
using Bayes rule would learn the truth or not, however, for the main
theory discussed in this section, the results hold even when relaxing
this assumption as shown in \citet{Heidhues2018}.

A key notion in this setting is that of a self-defeating
equilibrium\footnote{This notion is an adaptation of the Berk-Nash equilibrium in \citet{Esponda2016} to this setting with only one agent}.
A \emph{self-defeating equilibrium} is a belief and action pair,
\((p, a^*)\), such that the agent's belief about his type is
misspecified, and the outcome generated by the action is consistent with
the misspecified belief. The agent's belief is said to be \emph{stable}
when this happens.

Given the choice that sustains the self-defeating equilibrium, they
observe the \emph{average output} given by
\[\bar{h}(a^*) =\frac{1}{T}\sum_t \left[h(\hat{\theta}, \hat{\omega}, a^*)+\varepsilon_t\right]\]
where \(\hat{\theta}\) and \(\hat{\omega}\) is the type-state pair on
which the misspecified belief places all probability. That is, the
agents belief is \(p(\hat{\theta}, \hat{\omega}) = 1\) and
\(p(\theta, \omega) = 0\) for all other values of the parameter.

Within this framework, I consider two nested theories of belief
updating: the first one is a dogmatic modeler adapted from
\citet{Heidhues2018}. The second one is a switcher, as in
\citet{Ba2023}. The dogmatic modeler can be seen as a switcher with an
infinitely sticky initial belief so the two theories are nested.
However, outside of the limiting case, they produce different
predictions about the agent's behavior. I discuss each of them
separately in what follows.

@@

A dogmatic agent is characterized by a learning model where he does not
update his beliefs about \(\theta\) and instead, holds a degenerate
belief that places all probability on \(\hat{\theta}\), which is
potentially misspecified. In this case, no matter how much evidence he
gathers against being of type \(\hat{\theta}\), he will not update the
beliefs about his type. Any discrepancies between the observed outcomes
and his believed type are incorporated using the Bayes rule to update
his beliefs about \(\omega\). \citet{Heidhues2018} show that, under
certain assumptions on the per-period
utility,\footnote{The assumptions are that $u$ is twice continuously differentiable with: (i)$u_{aa}<0$ and $u_a(\underline{a},  \theta, \omega)>0>u_a(\Bar{a}, \theta, \omega)$, (ii) $u_{\theta}, u_{\omega}>0$ and (iii) $u_{a\theta}<0$ and $u_{a\omega}>0$. Where $\underline{a}$ is the
minimal action and $\Bar{a}$ is the maximal one. The direction of the derivatives is a normalization
and the results would hold even when the signs are reversed.} a dogmatic
modeler will inevitably fall into a self-defeating equilibrium. The
equilibrium will be such that the outcomes he observes reinforce his
belief on \(\omega\) in such a way that as \(t\to\infty\) the agent will
be sure that the state is some \(\omega^{'}\) consistent with their
believed type and the observed data. In other words, he will be in a
self-defeating equilibrium with a stable belief that places probability
one on the incorrect parameters \((\hat{\theta}, \omega_{\infty})\)

The mechanism by which the dogmatic agent falls into the self-defeating
equilibrium is the following: suppose the agent holds the dogmatic and
misspecified belief that they are type \(\hat{\theta}>\theta^*\). For
any prior over \(\omega\), the agent will be disappointed by the
outcome. That is:
\[\frac{1}{T}\sum_t \left[{h}(\theta^*, \omega^*, a_t^*)+\varepsilon_t\right] <\mathbb{E_w} \left[h(\hat{\theta}, \omega, a_t^*)\right]\]
where the right-hand-side of the inequality represents the average
outcome that the agent expected given his beliefs about their type and
the state. The left hand side is what they observed.

There are two possible sources for the disappointment: the first is that
the realized state \(\omega^*\) is lower than what the agent expected,
\(\mathbb{E}[\omega]\); the second source is that the agent is of type
\(\theta^*\) and therefore, for all possible states, his gain will be
lower than what he expected when the variance of the error term is low
or \(T\) is large. That is, when \(\frac{1}{T}\sum_t\varepsilon_t\) is
close to zero.

Because the agent is dogmatic, he will not update his beliefs about
\(\theta\) and, as a consequence, will attribute the disappointment to
the state being lower than expected. He will continue to update in this
way until he converges to a belief about \(\omega\) that is stable. Such
a belief will explain the observed utility perfectly and allow the agent
to rationalize his dogmatic belief about \(\theta\). Under the
assumptions of \citet{Heidhues2018}, there is a unique value of
\(\omega\) at which the belief is stable, I will refer to such value as
\(\omega_\infty\). This mechanism is illustrated in Example 1.

\textbf{Example 1: } \emph{Set \(A = \Omega=[\underline{a}, \infty)\)
and consider a student (she) with intrinsic ability \(\theta^*\geq 0\)
who faces a grading procedure \(\omega^*\) that is unknown to her. The
student knows that a higher \(\omega^*\) is more likely to yield a
higher grade. In particular, she knows the grade is given by
\((\theta^*+a)\omega^*\).}

\emph{The student must choose an effort level \(a\), which determines
her grade. For whatever the chosen effort level, the student must pay a
cost \(c(a) = \frac{1}{2}a^2\). She repeats this process for infinitely
many periods. Assume also that the student's prior is such that
\(\mathbb{E}[\omega]= \omega^*\) and she is dogmatic about being of type
\(\hat{\theta}>\theta^*\).\footnote{The example is illustrated for an overconfident student but the results are symmetric for a dogmatic student who initially places probability one on some $\tilde{\theta}<\theta^*$.}
Therefore, the student's payoff in period \(t\) is given by }

@@

In particular, for an agent that starts with a dogmatic belief that they
are of type \(\hat{\theta}\) but is willing to consider the alternative
explanation that they are of type \(\tilde{\theta}\), the agent will
switch to the alternative if

@@

In order to compare the predictions of the theories discussed above, I
develop a unifying example where the forces behind each of the theories
can be isolated. The example is modified from \citet{Heidhues2018} and
is adapted to be implementable in the lab.

@@

\caption{Probability of success for each type $\theta$, gamble $\omega$, and effort level $e$}
\label{tab:DGP}
\end{table}
